---
title: "modsite Workflow: End-to-End RNA Modification Site Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{modsite Workflow: End-to-End RNA Modification Site Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

`modsite` provides a complete pipeline for analysing RNA modification sites
detected by methods such as PUMseq, GLORIseq, and LIMEseq. The workflow
consists of six stages:

1. **Preprocessing** – merge per-sample pileup files, filter low-quality sites,
   annotate with genomic regions and known modification databases.
2. **Reporting** – compute per-sample summary statistics and assess sequencing
   saturation.
3. **Differential analysis (simple)** – two-group site comparison using t-test
   or Wilcoxon rank-sum test.
4. **Differential analysis (GLMM)** – Beta-Binomial GLMM or hierarchical
   Bayesian GLMM for multi-covariate designs, with optional dispersion
   stabilisation.
5. **Feature aggregation** – collapse site-level p-values to gene/transcript
   level using the Cauchy Combination Test (CCT).
6. **Visualisation** – metagene profiles and sequence logos.

```{r load}
library(modsite)
```

---

## Stage 1 – Preprocessing

### 1.1 Merge per-sample pileup files

Each sample produces a tab-separated pileup file with columns `chrom`, `pos`,
`ref`, `depth`, and nucleotide count columns (`A`, `C`, `G`, `T`). For larger
cohorts, it is often more convenient to discover files programmatically and
then pass the file paths and condition labels to `new_merger()`.

```{r merger}
base_dir <- "data/pileups"

sample_files <- list.files(
  path = base_dir,
  pattern = "^genome\\.sites\\.tsv$",
  recursive = TRUE,
  full.names = TRUE
)
sample_files <- normalizePath(sample_files, winslash = "/", mustWork = FALSE)
sample_files <- sample_files[!grepl("NC", sample_files)]

conditions <- c(
  rep("ctrl", 20),
  rep("case", 19)
)
stopifnot(length(sample_files) == length(conditions))

m <- new_merger(
  sample_files = sample_files,
  condition = conditions,
  modification_method = "PUMseq",
  min_modification_rate = 0.05,
  min_depth = 10L,
  group_missing_threshold = 0.2,
  group_filter_strategy = "all"
)

merged_df <- merge_samples(m)
dim(merged_df)
head(merged_df[, 1:8])
```

### 1.2 Remove low-quality samples

Drop samples with a high fraction of missing sites. This updates the merger
object in place:

```{r filter-samples}
filter_samples(m, max_missing_rate = 0.8)
merged_df <- m$merged_data
merger_summary(m)$per_sample |> lapply(\(x) x$missing_pct)
```

### 1.2.1 Visual QC for the merged table

The merger module also provides several sample-level visual diagnostics:

```{r merger-visual-qc}
plot_mod_rate_distribution(m)
plot_sample_correlation(m)
plot_merger_pca(m, impute_scope = "global")
consistency <- plot_detection_consistency(m, show_score_bar = TRUE)
plot_site_overlap(m)
```

### 1.3 Annotate genomic regions

Use a GTF file to assign each site to genomic regions (CDS, UTR5, UTR3,
intron, and so on) together with gene-level metadata:

```{r annotate-genome}
annotator <- new_genomic_annotator(
  gtf_file = "/path/to/Homo_sapiens.GRCh38.113.gtf"
)
merged_ann <- annotate_genomic_regions(annotator, merged_df)
table(merged_ann$genomic_region)
```

### 1.4 Cross-reference known modification databases

Match sites against a known modification CSV file. The package ships with a
small RMBase example file that is convenient for demos and tests:

```{r annotate-mods}
mod_file <- system.file("extdata", "RMBase_v3_human.csv", package = "modsite")
mod_ann <- new_mod_annotator(mod_file)
merged_ann <- annotate_known_mods(mod_ann, merged_ann)
table(merged_ann$is_known_mod, useNA = "ifany")
```

### 1.5 Save the preprocessed table

```{r save-merged}
save_merged(m, "results/merged_sites.tsv")

# Keep the annotated table on the merger for downstream GLMM/CCT steps.
m$merged_data <- merged_ann

utils::write.table(
  merged_ann,
  file = "results/merged_annotated.tsv",
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)
```

---

## Stage 2 – Reporting

### 2.1 Per-sample summary statistics

```{r sample-stats}
ss <- new_sample_stats(merged_ann)

# Overall per-sample summary
sample_summary_stats(ss)

# Per-chromosome breakdown
head(chromosome_stats(ss))

# Per-genomic-region breakdown
genomic_region_stats(ss)

# Modification rate distribution
modification_rate_bin_stats(ss)
```

### 2.2 Sequencing saturation analysis

Requires a BAM file and Bioconductor packages `Rsamtools` and
`GenomicAlignments` (listed in `Suggests`):

```{r saturation}
az <- new_saturation_analyzer(
  "data/ctrl1.bam",
  min_depth = 10L,
  min_mod_rate = 0.05,
  test_mode = FALSE
)

site_stats <- extract_site_stats(az)
head(site_stats)

sim <- run_saturation_analysis(
  az,
  random_seed = 42L,
  output_file = "results/saturation_ctrl1.png"
)
head(sim)
```

---

## Stage 3 – Differential analysis (simple two-group test)

### 3.1 Construct the analyzer

```{r diff-sites-new}
ctrl_cols <- unname(m$sample_names[m$condition == "ctrl"])
case_cols <- unname(m$sample_names[m$condition == "case"])

ds <- new_diff_sites(
  annotated_df    = merged_ann,
  group1_samples  = case_cols,
  group2_samples  = ctrl_cols,
  group1_name     = "Case",
  group2_name     = "Control",
  test_method     = "wilcox-test",
  p_value_threshold = 0.05,
  min_abs_log2fc  = 1.0,
  min_samples_per_group = 2L
)
```

### 3.2 Run the test and inspect results

```{r diff-sites-run}
results <- run_diff_sites(ds)
results <- results[order(results$adj.p.value), ]

# Top significant sites
sig <- results[results$significant == TRUE, ]
head(sig[order(sig$adj.p.value), c("chrom","pos","ref",
                                   "Case_mean","Control_mean",
                                   "log2fc","adj.p.value")])
```

### 3.3 Save results

```{r diff-sites-save}
save_diff_results(results, "results/diff_sites_wilcox.tsv")
save_diff_results(results, "results/all_sites_wilcox.tsv", include_all = TRUE)
```

---

## Stage 4 – Differential analysis (Beta-Binomial GLMM)

The GLMM approach supports multiple fixed effects, random effects, and
optional dispersion stabilisation via a weighted-least-squares phi trend.
It requires the `glmmTMB` and `broom.mixed` packages (listed in `Suggests`).
If you prefer a hierarchical Bayesian model with the same top-level output
layout, use `run_glmm_bayes()` instead of `run_glmm()`.

### 4.1 Attach sample metadata to the merger

```{r glmm-meta}
sample_ids <- unname(m$sample_names)

m$sample_meta <- data.frame(
  sample_id = sample_ids,
  group = factor(m$condition, levels = c("ctrl", "case")),
  batch = factor(rep(paste0("batch", 1:4), length.out = length(sample_ids))),
  stringsAsFactors = FALSE
)
```

### 4.2 (Optional) Estimate dispersion trend

Stabilise per-site over-dispersion by fitting a WLS regression of
site-level dispersion estimates onto site covariates. In practice, `motif`
and `strand` are often good starting points when those columns are available:

```{r phi-trend}
phi_vars <- intersect(c("motif", "strand"), colnames(m$merged_data))
stopifnot(length(phi_vars) > 0)

estimate_phi_trend(
  merger = m,
  phi_vars = phi_vars,
  min_depth_site = 5L,
  min_samples_per_site = 6L,
  verbose = TRUE
)
str(m$phi_trend)
```

### 4.3 Run the GLMM

```{r glmm-run}
glmm_out <- run_glmm(
  merger = m,
  fixed = c("group", "batch"),
  primary_term = "group",
  min_depth_site = 5L,
  min_samples_per_site = 4L,
  n_cores = 1L,
  on_error = "warn",
  phi = TRUE
)

# Primary-term results (one row per site)
ptb <- glmm_out$primary_term_backfill
ptb <- ptb[order(ptb$group_adj_p.value), ]
head(ptb, 10)

# Convergence summary
cat("Sites fitted OK:", sum(ptb$fit_ok, na.rm = TRUE), "/", nrow(ptb), "\n")
table(ptb$or_extreme, useNA = "ifany")
```

### 4.4 Run the hierarchical Bayesian GLMM

`run_glmm_bayes()` accepts the same merger-centered inputs and returns the
same top-level slots as `run_glmm()`. The difference is in the fitting
backend: site-level MAP estimation with empirical-Bayes hyperparameter
updates and Laplace-approximate posterior summaries.

```{r glmm-bayes-run}
glmm_bayes_out <- run_glmm_bayes(
  merger = m,
  fixed = c("group", "batch"),
  primary_term = "group",
  min_depth_site = 5L,
  min_samples_per_site = 4L,
  on_error = "warn",
  max_iter = 8L,
  verbose = TRUE
)

# Same top-level output structure as run_glmm()
bayes_ptb <- glmm_bayes_out$primary_term_backfill
head(bayes_ptb[order(bayes_ptb$group_adj_p.value), ])

# Bayesian metadata are stored in glmm_slot
glmm_bayes_out$glmm_slot$method
glmm_bayes_out$glmm_slot$converged
```

---

## Stage 5 – Feature-level aggregation (CCT)

Collapse site-level p-values to the gene level using the Cauchy Combination
Test:

```{r cct}
gene_res <- aggregate_feature_cct(
  site_df     = glmm_out$primary_term_backfill,
  feature_col = "gene_id",
  p_col       = "group_p.value",
  effect_col  = "group_est_logodds",
  weight_col  = NULL,
  min_sites   = 2L,
  adj_method  = "BH"
)

# Top differentially modified genes
head(gene_res[order(gene_res$cct_adj.p.value),
              c("gene_id","n_sites_used","cct_p.value",
                "cct_adj.p.value","mean_effect")])
```

---

## Stage 6 – Visualisation

### 6.1 Metagene profile

Requires the `annotator` and `merged_ann` objects created above:

```{r metagene}
mg <- new_metagene_analyzer(
  annotator = annotator,
  annotated_df = merged_ann,
  n_bins = 50L
)

mg <- calc_metagene_profile(mg, smooth = TRUE, span = 0.3)
plot_metagene(mg, title = "Metagene site density")

selected_samples <- c(ctrl_cols[1], case_cols[1])
mg_weighted <- calc_metagene_profile(
  mg,
  weight_cols = selected_samples,
  smooth = TRUE,
  span = 0.3
)
plot_metagene(
  mg_weighted,
  series_to_plot = c("count", selected_samples),
  title = "Metagene weighted signals"
)
```

### 6.2 Sequence logo

Visualise the sequence context around modification sites:

```{r seqlogo}
# Extract +/- 5 nt context sequences from the annotated table
sequences <- merged_ann$motif   # character vector of context sequences

ggseqlogo(sequences, method = "prob") +
  ggplot2::theme_classic() +
  ggplot2::ggtitle("Sequence context around modification sites")
```

---

## Session information

```{r session, eval = TRUE}
utils::sessionInfo()
```
