---
title: "modsite Workflow: End-to-End RNA Modification Site Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{modsite Workflow: End-to-End RNA Modification Site Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

`modsite` provides a complete pipeline for analysing RNA modification sites
detected by methods such as PUMseq, GLORIseq, and LIMEseq. The workflow
consists of six stages:

1. **Preprocessing** – merge per-sample pileup files, filter low-quality sites,
   annotate with genomic regions and known modification databases.
2. **Reporting** – compute per-sample summary statistics and assess sequencing
   saturation.
3. **Differential analysis (simple)** – two-group site comparison using t-test
   or Wilcoxon rank-sum test.
4. **Differential analysis (GLMM)** – Beta-Binomial GLMM for multi-covariate
   designs with optional dispersion stabilisation.
5. **Feature aggregation** – collapse site-level p-values to gene/transcript
   level using the Cauchy Combination Test (CCT).
6. **Visualisation** – metagene profiles and sequence logos.

```{r load}
library(modsite)
```

---

## Stage 1 – Preprocessing

### 1.1 Merge per-sample pileup files

Each sample produces a tab-separated pileup file with columns `chrom`, `pos`,
`ref`, `depth`, and nucleotide count columns (`A`, `C`, `G`, `T`).  Pass all
file paths along with their experimental conditions to `new_merger()`, then
call `merge_samples()` to produce the merged wide-format table.

```{r merger}
sample_files <- c(
  "data/ctrl1.tsv", "data/ctrl2.tsv", "data/ctrl3.tsv",
  "data/trt1.tsv",  "data/trt2.tsv",  "data/trt3.tsv"
)
conditions <- c("ctrl", "ctrl", "ctrl", "trt", "trt", "trt")

m <- new_merger(
  sample_files        = sample_files,
  condition           = conditions,
  modification_method = "PUMseq",
  min_modification_rate = 0.05,
  min_depth           = 10L
)

merged_df <- merge_samples(m)
dim(merged_df)
head(merged_df[, 1:6])
```

### 1.2 Remove low-quality samples

Drop samples with a high fraction of missing sites:

```{r filter-samples}
filter_samples(m, max_missing_rate = 0.4)
merger_summary(m)$per_sample |> lapply(\(x) x$missing_pct)
```

### 1.3 Annotate genomic regions

Use a GTF file to assign each site to a genomic region (CDS, UTR5, UTR3,
intron, etc.) and gene metadata:

```{r annotate-genome}
annotator <- new_genomic_annotator(
  gtf_file   = "ref/gencode.v47.annotation.gtf",
  genome_ver = "GRCh38"
)
merged_ann <- annotate_genomic_regions(annotator, merged_df)
table(merged_ann$genomic_region)
```

### 1.4 Cross-reference known modification databases

Match sites against a known modification BED/TSV file (e.g. m6A atlas):

```{r annotate-mods}
mod_ann <- new_mod_annotator(
  mod_file  = "ref/m6A_atlas.bed",
  mod_type  = "m6A",
  tolerance = 2L
)
merged_ann <- annotate_known_mods(mod_ann, merged_ann)
table(merged_ann$is_known_mod)
```

### 1.5 Save the preprocessed table

```{r save-merged}
save_merged(m, "results/merged_annotated.tsv")
```

---

## Stage 2 – Reporting

### 2.1 Per-sample summary statistics

```{r sample-stats}
sample_cols <- grep("^depth_", colnames(merged_ann), invert = TRUE,
                    value = TRUE) |>
  setdiff(c("chrom", "pos", "ref", "strand", "motif",
             "genomic_region", "gene_id", "gene_name", "gene_type",
             "is_known_mod"))

ss <- new_sample_stats(merged_ann, sample_cols = sample_cols)

# Overall per-sample summary
sample_summary_stats(ss)

# Per-chromosome breakdown
head(chromosome_stats(ss))

# Per-genomic-region breakdown
genomic_region_stats(ss)

# Modification rate distribution
modification_rate_bin_stats(ss)
```

### 2.2 Sequencing saturation analysis

Requires a BAM file and Bioconductor packages `Rsamtools` and
`GenomicAlignments` (listed in `Suggests`):

```{r saturation}
az  <- new_saturation_analyzer("data/ctrl1.bam", min_depth = 10L,
                               min_mod_rate = 0.05, test_mode = FALSE)
run_saturation_analysis(az, random_seed = 42L,
                        output_file = "results/saturation_ctrl1.png")
```

---

## Stage 3 – Differential analysis (simple two-group test)

### 3.1 Construct the analyzer

```{r diff-sites-new}
ctrl_cols <- paste0("ctrl", 1:3)
trt_cols  <- paste0("trt",  1:3)

ds <- new_diff_sites(
  annotated_df    = merged_ann,
  group1_samples  = trt_cols,
  group2_samples  = ctrl_cols,
  group1_name     = "Treatment",
  group2_name     = "Control",
  test_method     = "t-test",
  p_value_threshold = 0.05,
  min_abs_log2fc  = 1.0
)
```

### 3.2 Run the test and inspect results

```{r diff-sites-run}
results <- run_diff_sites(ds)

# Top significant sites
sig <- results[results$significant == TRUE, ]
head(sig[order(sig$adj.p.value), c("chrom","pos","ref",
                                   "Treatment_mean","Control_mean",
                                   "log2fc","adj.p.value")])
```

### 3.3 Save results

```{r diff-sites-save}
save_diff_results(results, "results/diff_sites_ttest.tsv")
save_diff_results(results, "results/all_sites_ttest.tsv", include_all = TRUE)
```

---

## Stage 4 – Differential analysis (Beta-Binomial GLMM)

The GLMM approach supports multiple fixed effects, random effects, and
optional dispersion stabilisation via a weighted-least-squares phi trend.
It requires the `glmmTMB` and `broom.mixed` packages (listed in `Suggests`).

### 4.1 Attach sample metadata to the merger

```{r glmm-meta}
m$sample_meta <- data.frame(
  sample_id = m$sample_names,
  group     = factor(conditions),
  stringsAsFactors = FALSE
)
```

### 4.2 (Optional) Estimate dispersion trend

Stabilise per-site over-dispersion by fitting a WLS regression of
site-level dispersion estimates onto site covariates:

```{r phi-trend}
estimate_phi_trend(m, phi_vars = "genomic_region", verbose = TRUE)
str(m$phi_trend)
```

### 4.3 Run the GLMM

```{r glmm-run}
glmm_out <- run_glmm(
  merger            = m,
  fixed             = "group",
  primary_term      = "group",
  min_depth_site    = 5L,
  min_samples_per_site = 4L,
  on_error          = "na",
  phi               = TRUE       # use dispersion trend from step 4.2
)

# Primary-term results (one row per site)
ptb <- glmm_out$primary_term_backfill
head(ptb[order(ptb$group_adj_p.value), ])

# Convergence summary
cat("Sites fitted OK:", sum(ptb$fit_ok), "/", nrow(ptb), "\n")
```

---

## Stage 5 – Feature-level aggregation (CCT)

Collapse site-level p-values to the gene level using the Cauchy Combination
Test:

```{r cct}
gene_res <- aggregate_feature_cct(
  site_df     = glmm_out$primary_term_backfill,
  feature_col = "gene_id",
  p_col       = "group_p.value",
  effect_col  = "group_est_logodds",
  weight_col  = NULL,
  min_sites   = 2L,
  adj_method  = "BH"
)

# Top differentially modified genes
head(gene_res[order(gene_res$cct_adj.p.value),
              c("gene_id","n_sites_used","cct_p.value",
                "cct_adj.p.value","mean_effect")])
```

---

## Stage 6 – Visualisation

### 6.1 Metagene profile

Requires a GTF/GFF file and a merged site table:

```{r metagene}
mg <- new_metagene_analyzer(
  gtf_file    = "ref/gencode.v47.annotation.gtf",
  sample_cols = trt_cols,
  n_bins      = 50L
)

profile <- calc_metagene_profile(mg, merged_ann)
plot_metagene(profile, title = "m6A metagene – PUMseq")
```

### 6.2 Sequence logo

Visualise the sequence context around modification sites:

```{r seqlogo}
# Extract +/- 5 nt context sequences from the annotated table
sequences <- merged_ann$motif   # character vector of context sequences

ggseqlogo(sequences, method = "prob") +
  ggplot2::theme_classic() +
  ggplot2::ggtitle("Sequence context around modification sites")
```

---

## Session information

```{r session, eval = TRUE}
utils::sessionInfo()
```
